\pkg{evtree}: Evolutionary Learning of Globally Optimal Classification and Regression Trees in \proglang{R}

### 2012-03-30 pkg/evtree/Submission_JSS/Revisions/response_revision_14032012.tex

We cleaned-up the notation and wording according to the reviewer's recommendations.
\\ \\

\textit{In the end, the authors gave honest comparisons of their methods to the traditional rpart and ctree methods, but it is unclear as to when and why evtree outperforms its competitors apart from the specific chessboard example given in Section 5.2.}

We added the following paragraph to the ``Conclusions'', as an explanation as to when and why evtree outperforms its competitors:

No obvious relationship between the data characteristics and evtree's
relative performance could be observed. Probably, one explanation is that for some datasets,
the recursive partitioning algorithms already yield (nearly) optimal trees, leaving little room for further
improvements. For other datasets, evtree's global search strategy can find more
complex interactions, resulting in better predictive accuracy and/or sparser trees.
\\ \\
\textbf{2 Spelling/Grammatical Errors}

@@

\textit{
\begin{enumerate}
\item (p.3) Above equation (2), ``predicition" should be ``prediction".
\item (p.7) First paragraph, midway, ``spit points" should be ``split points"
\item (p.8) Minor split rule mutation paragraph: ``less then" should be ``less than", and ``the split
point is change to to..." should be ``the split point is changed to".
\item (p.8) Crossover paragraph, last sentence: ``nods" should be ``nodes".
\item (p.8) Evaluation function first paragraph: should be ``The evaluation function represents the
requirements to which the population should adapt."
\item (p.14) Middle of first pararaph: ``bookclub" should be ``book club".
\end{enumerate}
}

We changed all spelling- and grammatical errors as suggested.
\\  \\
\textbf{3 Notation}

\textit{In Section 2, the authors introduce some notation for the tree setup, but the notation seems inconsistent in the paper. The authors write $\theta=(v_{n_1}, s_{n_1},\dots,v_{n_{M-1}},s_{n_{M-1}})$, where $n_r \in \{1,\dots, M_{max}-1\}$ but then also write $v_r$ and $s_r$ where $r=1,\dots,M-1$. Further, in Section 2.2, the authors talk about $s_r$ and $v_r$ where $r \in \{n_1, \dots, n_{M-1} \}$, which is immediately different from the indexing given
in equation (3). The authors should use a consistent notation for indexing the $v$ and $s$.}

We cleaned up the notation and made it consistent. A binary tree is now denoted as $\theta=(v_1, s_1, \dots, v_{M-1}, s_{M-1})$, were $r \in \{1,...,M-1\}$ are the internal \emph{nodes}, $v_r \in \{1, \dots ,P\}$ the associated splitting \emph{variables}, and $s_r$ the associated \emph{split} rules ($r = 1, \dots, M-1$). Now, in order to simplify the notation, the positions of the internal nodes are no longer made explicit. 
\\  \\
\textbf{4 Case Study}

\textit{The authors have provided very minimal information on the dataset. (p.14) The authors write
``Thus, the evolutionary tree uses a different cutoff in art for book club members that joined in the last year as opposed to older customers." The text does not explain which variable corresponds to ``join date", but based on Figure 3, one can indirectly infer that this variable is first. The authors have described te variable first as the number of months since the first purchase, but do not say that members join the book club upon making their first purchase.}

We added a table (Table 2), which shortly describes the variables of the Bookbinder's Book Club dataset. We assumed the join date to be equal to the data of the first purchase. To avoid confusions we changed the sentence:

@@

\textit{\item (p.7) Prune paragraph: We might append to the end of the sentence for clarity: ``...and no pruning occurs."}

We changed the sentence according to the recommendations.

\textit{\item (p.8) Minor split rule mutation paragraph: The description of the mutation rules is rather complex, with really 4 different cases to consider (nominal/$<$ 20 values, nominal/$\ge$ 20 values, continuous/$<$ 20 values, continuous/$\ge$ 20 values). It might be good to present this set of rules in a tabular format.}

The rules are now presented in an easy to read tabular format.

### 2012-03-30 pkg/evtree/inst/doc/evtree.Rnw
continued revision; reply is added to new folder 'Revisions'iin folder 'Submission_jss'
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@286 255f4249-2263-4417-b1ea-8498779e27a7

The main challenge in growing globally optimal trees is that the search
space is typically huge rendering full-grid searches computationally infeasible.
One possibility to solve this problem is to use stochastic
optimization methods like evolutionary algorithms. 
In practice, however, such stochastic methods
are rarely used in decision tree induction. One reason is probably that they are
computationally much more demanding than a recursive forward search but another one is
likely to be the lack of availability in major software packages.
In particular, while there are several packages for \proglang{R} \citep{team2011r} providing
forward-search tree algorithms, there is only little support for globally
optimal trees. The former group of packages includes (among others)
\pkg{rpart} \citep{therneau1997introduction}, the open-source
implementation of the CART algorithm; \pkg{party}, containing two tree algorithms
with unbiased variable selection and statistical stopping criteria
\citep{hothorn2006urp,zeileis2008mob}; and \pkg{RWeka} \citep{rweka}, the
\proglang{R} interface to \pkg{Weka} \citep{weka} with open-source implementations
of tree algorithms such as C4.5 or \textit{M5} \citep{quinlan1992lcc}.
A notable exception is the
\pkg{LogicReg} package \citep{logicreg} for logic regression, an algorithm for globally optimal
trees based on binary covariates only  and using simulated annealing. See
\cite{hothorn2011ctv} for an overview of further recursive partitioning packages
for \proglang{R}.

@@

To show why forward-search recursive partitioning algorithms typically
lead to globablly suboptimal solutions, their parameter spaces and optimization
problems are presented and contrasted in a unified notation. Although all
arguments hold more generally, only binary tree models with some maximum number
of terminal nodes $M_{\max}$ are considered. Both restrictions make the notation somewhat
simpler while not really restricting the problem: (a)~Multiway splits are equivalent
to a sequence of binary splits in predictions and number of resulting subsamples.
(b)~The maximal size of the tree is always limited by the number of observations
in the learning sample.

In the following, a binary tree model with $M$ terminal nodes (which consequently has
$M-1$ internal splits) is denoted by
%
\begin{equation}
  \theta ~=~ (v_{1}, s_{1},...,v_{M-1},s_{M-1} ), \label{eq:tree}
\end{equation}
%
where $r \in \{1,...,M_{\max}-1\}$ are the internal \emph{nodes}, $v_r \in \{1, \dots ,P\}$ the associated splitting
\emph{variables}, and $s_r$ the associated \emph{split} rules ($r = 1, \dots, M-1$).
Depending on the domain of $X_{v_r}$, the split rule $s_r$ contains either a cutoff
(for ordered and numeric variables) or a nonempty subset of $\{1, \dots, c\}$
(for a categorical variable with $c$~levels), determining which observations are
sent to the first or second subsample. In the former case, there are $u-1$ possible
split rules if $X_{v_r}$ takes $u$ distinct values; and in the latter case, there
are $2^{c-1} - 1$ possible splits. Thus, the product of all of these combinations
forms all potential elements $\theta$ from $\Theta_M$, the space of conceivable
trees with $M$ terminal nodes. The overall parameter space is then 
$\Theta =\bigcup_{M=1}^{M_{\max}} \Theta_{M}$ (which in practice is often reduced
by exluding elements $\theta$ resulting in too small subsamples etc.).

@@

As done by \cite{breiman1984cart}, let the complexity of a tree be measured by a
function of the number of terminal nodes, without further considering the depth
or the shape of trees. The goal is then to find that classification and regression tree
which optimizes some tradeoff between prediction performance and complexity:
%
\begin{equation} \label{eq:globally}
  \hat{\theta} ~=~ \argmin_{\theta \in \Theta} \; \mathrm{loss}\{Y, f(X,\theta)\} ~+~ \mathrm{comp}(\theta).
\end{equation}
%
where $\mathrm{loss}(\cdot, \cdot)$ is a suitable loss function for the
domain of $Y$; typically, the misclassification rate MC 
and the mean squared error MSE are employed for classification and regression,
respectively. The function $\mathrm{comp}(\cdot)$ is a function that is monotonically
non-decreasing in the number of terminal nodes $M$ of the tree $\theta$, thus penalizing
more complex models in the tree selection selection process.
Note that finding $\hat{\theta}$ requires a search over all $\Theta_M$.

@@

Note that the notation above uses an exhaustive search for the $r$-th split,
jointly over $(v_r, s_r)$, as is employed in CART or C4.5. So-called \emph{unbiased}
recursive partitioning techniques modify this search by first seleting the variable $v_r$ using
statistical significance tests and subsequently selecting the optimal split $s_r$ for that
particular variable. This approach is used in conditional inference trees
\citep[see][for references to other algorithms]{hothorn2006urp} and avoids selecting
variables with many potential splits more often than those with fewer potential
splits.

@@

Another class of algorithms is given by stochastic optimization methods that,
given an initial tree, seek improved solutions through stochastic changes to
the tree structure. Thus, these algorithms try to explore the full parameter
space $\Theta$ but cannot be guaranteed to find the globally optimal solution
but only an approximation thereof. Besides evolutionary algorithms
\citep{koza1991concept}, \textit{Bayesian CART} \citep{denison1998bca} and
\textit{simulated annealing} \citep{sutton1992ict} were used successfully to
solve difficult classification and regression tree problems.
\cite{koza1991concept} first formulated the concept of using evolutionary
algorithms as a stochastic optimization method to build classification and
regression trees. \cite{papagelis2001bdt} presented a classification tree
algorithm and provided results on several datasets from the UCI machine learning
repository \citep{Frank+Asuncion:2010}. Another method for the construction of
classification and regression trees via evolutionary algorithms was introduced by
\cite{gray2008cta} and \cite{fan2005rta}, respectively.
\cite{cantu2003inducing} used an evolutionary algorithm to induce so-called oblique
classification trees.

@@

Each tree of the population is initialized with a valid, randomly generated,
split rule in the root node. First, $v_1$ is selected with uniform probability
from $1,...,P$. Second, a split point $s_1$ is selected. If $X_{v_1}$ is numeric
or ordinal with $u$ unique values, a split point $s_1$ is selected with uniform
probability from the $u-1$ possible split points of $X_{v_1}$. If $X_{v_1}$ is
nominal and has $c$ categories, each $k=1,...,c$ has a probability of $50\%$ to
be assigned to the left or the right daughter node. In cases where all $k$ are
allocated to the same terminal node, one of the $c$ categories is allocated to 
the other terminal node, to have the effect of ensuring both terminal nodes are nonempty.
If this procedure results in a non-valid split rule, the two steps of random split variable 
selection and split point selection are repeated. With the definition of $r=1$ and the 
selection of $v_1$ and $s_1$, the initialization is complete and each individual of
the population of trees is of type ${\theta}_{1}= (v_1, s_1)$.

@@

\textit{Minor split rule mutation} is similar to the \textit{major split rule
mutation} operator. However, it does not alter $v_r$ and only changes the split point
$s_r$ by a minor degree, which is defined by one of the following 4 cases:
\begin{itemize}
\item {$X_{v_r}$ is numerical or ordinal and has at least $20$ unique values:}
The split point $s_r$ is randomly shifted by a non-zero number of unique values of $X_{v_r}$ 
that is not larger than $10\%$ of the range of unique values.
\item {$X_{v_r}$ is numerical or ordinal and has less than $20$ unique values:}
The split point is changed to the next larger, or the next lower, unique value of $X_{v_r}$.
\item {$X_{v_r}$ is nominal and has at least $20$ categories:}  at least one and at most $10\%$ of the variable's
categories are changed.
\item {$X_{v_r}$ is nominal and has less than $20$ categories:} One of the categories is randomly
modified.
\end{itemize}
In cases where subsequent nodes become invalid, further
split points are searched that preserve the tree's topology. After five
non-successful attempts at finding a topology preserving split point, the
non-valid nodes are pruned.

@@

The principal function of the \pkg{evtree} package is the eponymous function
\code{evtree()} taking arguments
%
\begin{Code}
  evtree(formula, data = list(), weights = NULL, subset = NULL,
    control = evtree.control(...), ...)
\end{Code}
%
where \code{formula}, \code{data}, \code{weights}, and \code{subset} specify
the data in the usual way, e.g., via \code{formula = y ~ x1 + x2}.
Additionally, \code{control} comprises a list of control parameters
%
\begin{Code}
  evtree.control(minbucket = 7L, minsplit = 20L, maxdepth = 9L, 
    niterations = 10000L, ntrees = 100L, alpha = 1,
    operatorprob = list(pmutatemajor = 0.2, pmutateminor = 0.2,
      pcrossover = 0.2, psplit = 0.2, pprune = 0.2),
    seed = NULL, ...)
\end{Code}
%
where the parameters \code{minbucket}, \code{minsplit}, and \code{maxdepth}
constrain the solution to a minimum number of observations in each terminal
node, a minimum number of observation in each internal node and a maximum tree
depth. Note that the memory requirements increase by the square of the maximum
tree depth. Parameter \code{alpha} regulates the complexity parameter
$\alpha$ in Equation~\ref{eq:classification} and~\ref{eq:regression}, respectively.
\code{niterations} and \code{ntrees} specify the maximum number of iterations
and the number of trees in the population, respectively. With the argument
\code{operatorprob}, user-specified probabilities for the variation operators
can be defined. For making computations reproducible, argument \code{seed} is
an optional integer seed for the random number generator (at \proglang{C++} level).
If not specified, the random number generator is initialized by \verb|as.integer(runif(1, max = 2^16))|
in order to inherit the state of \code{.Random.seed} (at \proglang{R} level).
If set to \code{-1L}, the seed is initializied by the system time.

@@

All trees are constrained to have a minimum number of 10~observations per terminal node.
Additionally, a significance level of $1\%$ is employed in the construction of
conditional inference trees which is more appropriate than the default 5\%~level for
$1,300$~observations. To provide uniform visualizations and predictions of the
fitted models, `\code{party}' objects are used to represent all trees. For, `\code{rpart}'
trees \pkg{partykit} provides a suitable \code{as.party()} method while a
reimplementation of \code{ctree()} is provided in \pkg{partykit} (as opposed to
the original in \pkg{party}) that directly leverages the `\code{party}' infrastructure.

First, the data is loaded and the forward-search trees are grown
with and without depth restriction, visualizing the unrestricted trees
in Figure~\ref{fig:example_rpart_ctree}.
%
<<BBBClub-rpart-ctree, echo=TRUE, eval=FALSE>>=
data("BBBClub", package = "evtree")
library("rpart")
rp  <- as.party(rpart(choice ~ ., data = BBBClub, minbucket = 10))
rp2 <- as.party(rpart(choice ~ ., data = BBBClub, minbucket = 10,
  maxdepth = 2))
ct  <- ctree(choice ~ ., data = BBBClub, minbucket = 10, mincrit = 0.99)
ct2 <- ctree(choice ~ ., data = BBBClub, minbucket = 10, mincrit = 0.99,
  maxdepth = 2)
plot(rp)
plot(ct)
@
%
With the objective of building a smaller, but at still accurate tree,
\code{evtree} is constrained to a maximum tree depth of $2$,
see Figure~\ref{fig:example_evtree}.
%
<<BBBClub-evtree, echo=TRUE, eval=FALSE>>=
set.seed(1090)
ev <- evtree(choice ~ ., data = BBBClub, minbucket = 10, maxdepth = 2)
@
%
<<BBBClub-cache, echo=FALSE, eval=TRUE>>=
if(cache & file.exists("BBBClub-trees.rda")) {
load("BBBClub-trees.rda")
} else {
<<BBBClub-rpart-ctree>>
<<BBBClub-evtree>>
if(cache) {
  save(rp, rp2, ct, ct2, ev, file = "BBBClub-trees.rda")
} else {
  if(file.exists("BBBClub-trees.rda")) file.remove("BBBClub-trees.rda")
}
}
@
%
\setkeys{Gin}{width=1\textwidth}
\begin{figure}[p!]
\centering
\begin{minipage}{1\textwidth}
\begin{flushleft} \code{rpart} \end{flushleft}
\centering
\setkeys{Gin}{width=0.75\textwidth}
<<BBBClub-rpart-plot, fig=TRUE, echo=FALSE, width=8.1, height=7>>=
plot(rp)
@
\end{minipage}
\begin{minipage}{1\textwidth}
\begin{flushleft} \code{ctree} \end{flushleft}
\centering
\setkeys{Gin}{width=1\textwidth}
<<BBBClub-ctree-plot, fig=TRUE, echo=FALSE,  width=11.4, height=5.9>>=
plot(ct)
@
\end{minipage}
\caption{\label{fig:example_rpart_ctree} Trees for customer targeting constructed by \code{rpart}
  (upper panel) and \code{ctree} (lower panel). The target variable is the customer's choice of buying the book.
  The variables used for splitting are the number of art books purchased previously (\code{art}), the number
  of months since the first purchase (\code{first}), the frequency of previous purchases at the Bookbinder's Book Club
  (\code{freq}) and the customer's \code{gender}.}
\end{figure}
%
The resulting evolutionary tree is printed below and visualized in
Figure~\ref{fig:example_evtree}.
%
<<BBBClub-evtree-display>>=
plot(ev)
ev
@
%
\setkeys{Gin}{width=1\textwidth}
\begin{figure}[t!]
\begin{flushleft} \code{evtree} \end{flushleft}
\centering
\setkeys{Gin}{width=0.75\textwidth} 
<<BBBClub-evtree-plot, fig=TRUE, echo=FALSE, width=7.5, height=4.9>>=
plot(ev)
@
\caption{\label{fig:example_evtree} Tree for customer targeting constructed by \code{evtree}
   The target variable is the customer's choice of buying the book. The variables used for splitting are
   the number of art books purchased previously (\code{art}) and the number of months since the first purchase
    \code{first}.}
\end{figure}

Not surprisingly, the explanatory variable \code{art} -- the number of art books purchased previously at the
book club -- plays a key role in all constructed classification trees along with the number
of months since the first purchase (\code{first}), the frequency of previous purchases
(\code{freq}) and the customer's \code{gender}. Interestingly, though, the forward-search trees select
the arguably most important variable in the first split while the evolutionary tree uses \code{first}
in the first split and \code{art} in both second splits.
Thus, the evolutionary tree uses a different cutoff in \code{art} for book club members that bought there
first book in the last year as opposed to older customers.
While the former are predicted to be buyers if they previously bought
at least one art book, the latter are predicted to purchase the advertised art book only if they
previously bought at least two other art books. Certainly, this classification is easy to understand
and communicate (helped by Figure~\ref{fig:example_evtree}) to practitioners.

@@

Although the constructed trees are considerably different, the code above shows
that the predictive accuracy is rather similar. Moreover, below we see that the
structure of the individual predictions on the dataset are rather similar as well:
%
<<evtree-structure>>=
ftable(tab <- table(evtree = predict(ev), rpart  = predict(rp),
  ctree  = predict(ct), observed = BBBClub$choice))
sapply(c("evtree", "rpart", "ctree"), function(nam) {
  mt <- margin.table(tab, c(match(nam, names(dimnames(tab))), 4))
  c(abs = as.vector(rowSums(mt))[2],
    rel = round(100 * prop.table(mt, 1)[2, 2], digits = 3))
})
@
%
In this case, \code{evtree} classifies less
customers ($186$) as buyers as \code{rpart} ($216$) and \code{ctree} ($238$). However,
\code{evtree} achieves the highest proportion of correct classification among
the declared buyers: $72.6\%$ compared to $70.8\%$
(\code{rpart}) and $66.4\%$ (\code{ctree}).

@@

All models are constrained to a minimum number of 7~observations per terminal node,
20~observations per internal node and a maximum tree depth of~9. Apart from
that, the default settings of the algorithms are used. For assessment
of significant differences in predictive accuracy and complexity, respectively,
Dunnett's correction from \proglang{R}~package \pkg{multcomp} \citep{multcomp} was used
for calculating simultaneous $95\%$ confidence intervals on the individual datasets.

As missing values are currently not supported by \code{evtree} (e.g., by surrogate splits),
the 16~missing values in the \textit{Breast Cancer Database} -- the only dataset in the
study with missing values -- were removed before analysis.

>>>>

The main challenge in growing globally optimal trees is that the search
space is typically huge, rendering full-grid searches computationally infeasible.
One possibility to solve this problem is to use stochastic
optimization methods like evolutionary algorithms. 
In practice, however, such stochastic methods
are rarely used in decision tree induction. One reason is probably that they are
computationally much more demanding than a recursive forward search but another one is
likely to be the lack of availability in major software packages.
In particular, while there are several packages for \proglang{R} \citep{team2011r} providing
forward-search tree algorithms, there is only little support for globally
optimal trees. The former group of packages includes (among others)
\pkg{rpart} \citep{therneau1997introduction}, the open-source
implementation of the CART algorithm; \pkg{party}, containing two tree algorithms
with unbiased variable selection and statistical stopping criteria
\citep{hothorn2006urp,zeileis2008mob}; and \pkg{RWeka} \citep{rweka}, the
\proglang{R} interface to \pkg{Weka} \citep{weka} with open-source implementations
of tree algorithms such as C4.5 or \textit{M5} \citep{quinlan1992lcc}.
A notable exception is the
\pkg{LogicReg} package \citep{logicreg} for logic regression, an algorithm for globally optimal
trees based on binary covariates only  and using simulated annealing. See
\cite{hothorn2011ctv} for an overview of further recursive partitioning packages
for \proglang{R}.

@@

To show why forward-search recursive partitioning algorithms typically
lead to globablly suboptimal solutions, their parameter spaces and optimization
problems are presented and contrasted in a unified notation. Although all
arguments hold more generally, only binary tree models with some maximum number
of terminal nodes $M_{\max}$ are considered. Both restrictions make the notation somewhat
simpler while not really restricting the problem: (a)~Multiway splits are equivalent
to a sequence of binary splits in predictions and number of resulting subsamples;
(b)~the maximal size of the tree is always limited by the number of observations
in the learning sample.

In the following, a binary tree model with $M$ terminal nodes (which consequently has
$M-1$ internal splits) is denoted by
%
\begin{equation}
  \theta ~=~ (v_{1}, s_{1},...,v_{M-1},s_{M-1} ), \label{eq:tree}
\end{equation}
%
where $r \in \{1,...,M-1\}$ are the internal \emph{nodes}, $v_r \in \{1, \dots ,P\}$ the associated splitting
\emph{variables}, and $s_r$ the associated \emph{split} rules ($r = 1, \dots, M-1$).
Depending on the domain of $X_{v_r}$, the split rule $s_r$ contains either a cutoff
(for ordered and numeric variables) or a non-empty subset of $\{1, \dots, c\}$
(for a categorical variable with $c$~levels), determining which observations are
sent to the first or second subsample. In the former case, there are $u-1$ possible
split rules if $X_{v_r}$ takes $u$ distinct values; and in the latter case, there
are $2^{c-1} - 1$ possible splits. Thus, the product of all of these combinations
forms all potential elements $\theta$ from $\Theta_M$, the space of conceivable
trees with $M$ terminal nodes. The overall parameter space is then 
$\Theta =\bigcup_{M=1}^{M_{\max}} \Theta_{M}$ (which in practice is often reduced
by exluding elements $\theta$ resulting in too small subsamples etc.).

@@

As done by \cite{breiman1984cart}, let the complexity of a tree be measured by a
function of the number of terminal nodes, without further considering the depth
or the shape of trees. The goal is then to find that classification and regression tree
which optimizes some tradeoff between prediction performance and complexity:
%
\begin{equation} \label{eq:globally}
  \hat{\theta} ~=~ \argmin_{\theta \in \Theta} \; \mathrm{loss}\{Y, f(X,\theta)\} ~+~ \mathrm{comp}(\theta).
\end{equation}
%
where $\mathrm{loss}(\cdot, \cdot)$ is a suitable loss function for the
domain of $Y$; typically, the misclassification rate MC 
and the mean squared error MSE are employed for classification and regression,
respectively. The function $\mathrm{comp}(\cdot)$ is a function that is monotonically
non-decreasing in the number of terminal nodes $M$ of the tree $\theta$, thus penalizing
more complex models in the tree selection process.
Note that finding $\hat{\theta}$ requires a search over all $\Theta_M$.

@@

Note that the notation above uses an exhaustive search for the $r$-th split,
jointly over $(v_r, s_r)$, as is employed in CART or C4.5. So-called \emph{unbiased}
recursive partitioning techniques modify this search by first selecting the variable $v_r$ using
statistical significance tests and subsequently selecting the optimal split $s_r$ for that
particular variable. This approach is used in conditional inference trees
\citep[see][for references to other algorithms]{hothorn2006urp} and avoids selecting
variables with many potential splits more often than those with fewer potential
splits.

@@

Another class of algorithms is given by stochastic optimization methods that,
given an initial tree, seek improved solutions through stochastic changes to
the tree structure. Thus, these algorithms try to explore the full parameter
space $\Theta$ but cannot be guaranteed to find the globally optimal solution
but only an approximation thereof. Besides evolutionary algorithms
\citep{koza1991concept}, \textit{Bayesian CART} \citep{denison1998bca}, and
\textit{simulated annealing} \citep{sutton1992ict} were used successfully to
solve difficult classification and regression tree problems.
\cite{koza1991concept} first formulated the concept of using evolutionary
algorithms as a stochastic optimization method to build classification and
regression trees. \cite{papagelis2001bdt} presented a classification tree
algorithm and provided results on several datasets from the UCI machine learning
repository \citep{Frank+Asuncion:2010}. Another method for the construction of
classification and regression trees via evolutionary algorithms was introduced by
\cite{gray2008cta} and \cite{fan2005rta}, respectively.
\cite{cantu2003inducing} used an evolutionary algorithm to induce so-called oblique
classification trees.

@@

Each tree of the population is initialized with a valid, randomly generated,
split rule in the root node. First, $v_1$ is selected with uniform probability
from $1,...,P$. Second, a split point $s_1$ is selected. If $X_{v_1}$ is numeric
or ordinal with $u$ unique values, a split point $s_1$ is selected with uniform
probability from the $u-1$ possible split points of $X_{v_1}$. If $X_{v_1}$ is
nominal and has $c$ categories, each $k=1,...,c$ has a probability of $50\%$ to
be assigned to the left or the right daughter node. In cases where all $k$ are
allocated to the same terminal node, one of the $c$ categories is allocated to 
the other terminal node, to have the effect of ensuring both terminal nodes are nonempty.
If this procedure results in a non-valid split rule, the two steps of random split variable 
selection and split point selection are repeated. With the definition of $r=1$ (the root node) and the 
selection of $v_1$ and $s_1$, the initialization is complete and each individual of
the population of trees is of type ${\theta}_{1}= (v_1, s_1)$.

@@

\textit{Minor split rule mutation} is similar to the \textit{major split rule
mutation} operator. However, it does not alter $v_r$ and only changes the split point
$s_r$ by a minor degree, which is defined by one of the following 4 cases:
\begin{itemize}
\item {$X_{v_r}$ is numerical or ordinal and has at least $20$ unique values:}
The split point $s_r$ is randomly shifted by a non-zero number of unique values of $X_{v_r}$ 
that is not larger than $10\%$ of the range of unique values.
\item {$X_{v_r}$ is numerical or ordinal and has less than $20$ unique values:}
The split point is changed to the next larger, or the next lower, unique value of $X_{v_r}$.
\item {$X_{v_r}$ is nominal and has at least $20$ categories:} At least one and at most $10\%$ of the variable's
categories are changed.
\item {$X_{v_r}$ is nominal and has less than $20$ categories:} One of the categories is randomly
modified.
\end{itemize}
In cases where subsequent nodes become invalid, further
split points are searched that preserve the tree's topology. After five
non-successful attempts at finding a topology preserving split point, the
non-valid nodes are pruned.

@@

The principal function of the \pkg{evtree} package is the eponymous function
\code{evtree()} taking arguments
%
\begin{Code}
  evtree(formula, data = list(), weights = NULL, subset = NULL,
    control = evtree.control(...), ...)
\end{Code}
%
where \code{formula}, \code{data}, \code{weights}, and \code{subset} specify
the data in the usual way, e.g., via \code{formula = y ~ x1 + x2}.
Additionally, \code{control} comprises a list of control parameters
%
\begin{Code}
  evtree.control(minbucket = 7L, minsplit = 20L, maxdepth = 9L, 
    niterations = 10000L, ntrees = 100L, alpha = 1,
    operatorprob = list(pmutatemajor = 0.2, pmutateminor = 0.2,
      pcrossover = 0.2, psplit = 0.2, pprune = 0.2),
    seed = NULL, ...)
\end{Code}
%
where the parameters \code{minbucket}, \code{minsplit}, and \code{maxdepth}
constrain the solution to a minimum number of observations in each terminal
node, a minimum number of observations in each internal node, and a maximum tree
depth. Note that the memory requirements increase by the square of the maximum
tree depth. Parameter \code{alpha} regulates the complexity parameter
$\alpha$ in Equations~\ref{eq:classification} and~\ref{eq:regression}, respectively.
\code{niterations} and \code{ntrees} specify the maximum number of iterations
and the number of trees in the population, respectively. With the argument
\code{operatorprob}, user-specified probabilities for the variation operators
can be defined. For making computations reproducible, argument \code{seed} is
an optional integer seed for the random number generator (at \proglang{C++} level).
If not specified, the random number generator is initialized by \verb|as.integer(runif(1, max = 2^16))|
in order to inherit the state of \code{.Random.seed} (at \proglang{R} level).
If set to \code{-1L}, the seed is initializied by the system time.

@@

All trees are constrained to have a minimum number of 10~observations per terminal node.
Additionally, a significance level of $1\%$ is employed in the construction of
conditional inference trees, which is more appropriate than the default 5\%~level for
$1,300$~observations. To provide uniform visualizations and predictions of the
fitted models, `\code{party}' objects are used to represent all trees. For `\code{rpart}'
trees, \pkg{partykit} provides a suitable \code{as.party()} method while a
reimplementation of \code{ctree()} is provided in \pkg{partykit} (as opposed to
the original in \pkg{party}) that directly leverages the `\code{party}' infrastructure.

First, the data is loaded and the forward-search trees are grown
with and without depth restriction, visualizing the unrestricted trees
in Figure~\ref{fig:example_rpart_ctree}.
%
<<BBBClub-rpart-ctree, echo=TRUE, eval=FALSE>>=
data("BBBClub", package = "evtree")
library("rpart")
rp  <- as.party(rpart(choice ~ ., data = BBBClub, minbucket = 10))
rp2 <- as.party(rpart(choice ~ ., data = BBBClub, minbucket = 10,
  maxdepth = 2))
ct  <- ctree(choice ~ ., data = BBBClub, minbucket = 10, mincrit = 0.99)
ct2 <- ctree(choice ~ ., data = BBBClub, minbucket = 10, mincrit = 0.99,
  maxdepth = 2)
plot(rp)
plot(ct)
@
%
With the objective of building a smaller, but at still accurate tree,
\code{evtree} is constrained to a maximum tree depth of $2$,
see Figure~\ref{fig:example_evtree}.
%
<<BBBClub-evtree, echo=TRUE, eval=FALSE>>=
set.seed(1090)
ev <- evtree(choice ~ ., data = BBBClub, minbucket = 10, maxdepth = 2)
@
%
<<BBBClub-cache, echo=FALSE, eval=TRUE>>=
if(cache & file.exists("BBBClub-trees.rda")) {
load("BBBClub-trees.rda")
} else {
<<BBBClub-rpart-ctree>>
<<BBBClub-evtree>>
if(cache) {
  save(rp, rp2, ct, ct2, ev, file = "BBBClub-trees.rda")
} else {
  if(file.exists("BBBClub-trees.rda")) file.remove("BBBClub-trees.rda")
}
}
@
%
\setkeys{Gin}{width=1\textwidth}
\begin{figure}[p!]
\centering
\begin{minipage}{1\textwidth}
\begin{flushleft} \code{rpart} \end{flushleft}
\centering
\setkeys{Gin}{width=0.75\textwidth}
<<BBBClub-rpart-plot, fig=TRUE, echo=FALSE, width=8.1, height=7>>=
plot(rp)
@
\end{minipage}
\begin{minipage}{1\textwidth}
\begin{flushleft} \code{ctree} \end{flushleft}
\centering
\setkeys{Gin}{width=1\textwidth}
<<BBBClub-ctree-plot, fig=TRUE, echo=FALSE,  width=11.4, height=5.9>>=
plot(ct)
@
\end{minipage}
\caption{\label{fig:example_rpart_ctree} Trees for customer targeting constructed by \code{rpart}
  (upper panel) and \code{ctree} (lower panel). The target variable is the customer's choice of buying the book.
  The variables used for splitting are the number of art books purchased previously (\code{art}), the number
  of months since the first purchase (\code{first}), the frequency of previous purchases at the Bookbinder's Book Club (\code{freq}), and the customer's \code{gender}.}
\end{figure}
%
The resulting evolutionary tree is printed below and visualized in
Figure~\ref{fig:example_evtree}.
%
<<BBBClub-evtree-display>>=
plot(ev)
ev
@
%
\setkeys{Gin}{width=1\textwidth}
\begin{figure}[t!]
\begin{flushleft} \code{evtree} \end{flushleft}
\centering
\setkeys{Gin}{width=0.75\textwidth} 
<<BBBClub-evtree-plot, fig=TRUE, echo=FALSE, width=7.5, height=4.9>>=
plot(ev)
@
\caption{\label{fig:example_evtree} Tree for customer targeting constructed by \code{evtree}
   The target variable is the customer's choice of buying the book. The variables used for splitting are
   the number of art books purchased previously (\code{art}), and the number of months since the first purchase
    (\code{first}).}
\end{figure}

Not surprisingly, the explanatory variable \code{art} -- the number of art books purchased previously at the
book club -- plays a key role in all constructed classification trees along with the number
of months since the first purchase (\code{first}), the frequency of previous purchases
(\code{freq}), and the customer's \code{gender}. Interestingly, though, the forward-search trees select
the arguably most important variable in the first split while the evolutionary tree uses \code{first}
in the first split and \code{art} in both second splits.
Thus, the evolutionary tree uses a different cutoff in \code{art} for book club members that bought there
first book in the last year as opposed to older customers.
While the former are predicted to be buyers if they had previously bought
at least one art book, the latter are predicted to purchase the advertised art book only if they
had previously bought at least two other art books. Certainly, this classification is easy to understand
and communicate (helped by Figure~\ref{fig:example_evtree}) to practitioners.

@@

Although the constructed trees are considerably different, the code above shows
that the predictive accuracy is rather similar. Moreover, below we see that the
structure of the individual predictions on the dataset are rather similar as well:
%
<<evtree-structure>>=
ftable(tab <- table(evtree = predict(ev), rpart  = predict(rp),
  ctree  = predict(ct), observed = BBBClub$choice))
sapply(c("evtree", "rpart", "ctree"), function(nam) {
  mt <- margin.table(tab, c(match(nam, names(dimnames(tab))), 4))
  c(abs = as.vector(rowSums(mt))[2],
    rel = round(100 * prop.table(mt, 1)[2, 2], digits = 3))
})
@
%
In this case, \code{evtree} classifies fewer
customers ($186$) as buyers as \code{rpart} ($216$) and \code{ctree} ($238$). However,
\code{evtree} achieves the highest proportion of correct classification among
the declared buyers: $72.6\%$ compared to $70.8\%$
(\code{rpart}) and $66.4\%$ (\code{ctree}).

@@

All models are constrained to a minimum number of 7~observations per terminal node,
20~observations per internal node, and a maximum tree depth of~9. Apart from
that, the default settings of the algorithms are used. For the assessment
of significant differences in predictive accuracy and complexity, respectively,
Dunnett's correction from \proglang{R}~package \pkg{multcomp} \citep{multcomp} was used
for calculating simultaneous $95\%$ confidence intervals on the individual datasets.

As missing values are currently not supported by \code{evtree} (e.g., by surrogate splits),
the 16~missing values in the \textit{Breast cancer database} -- the only dataset in the
study with missing values -- were removed before analysis.
